---
title: "Gov 50 Final Project"
author: "Veronica Ruiz"
description: "Final Project on Parental Discrimination in Hiring"
output:
  distill::distill_article:
    self_contained: false
---



## Parental Discrimination
## By Veronica Ruiz

Introduction
Does being a parent hurt an applicant's probability of being hired? I hypothesize that being a parent can negatively affect an individual's chances of receiving a call back in the hiring process. Many occupations demand a large amount of time from their employees and prefer to hire individuals who can fully commit to their responsibilities. Stereotypical views of parenthood could have an effect on the hiring process. Since raising children is a life commitment - many occupations may have the view that individuals with children would be more unreliable and require more benefits - leading to an aversion to hiring parents. I also would like to see the difference across different genders to see if women - who are usually ascribed the role of primary caretaker - experience more discrimination as mothers than men as fathers.  
This type of hiring discrimination - if it is actually occuring - could harm many families and communities economically.


Data

The data for this research is from a cross sectional study found here(https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/4TTCEY).

The explanatory variable will be parent's gender and parental status. Parental status is measured as variable parent in this study and assigns a 0 to applicants without children and a 1 to individuals with children. The applications in the study signify if the applicant has children by mentioning children in their hobbies or by listing the PTA as one of their volunteering experiences. The outcome variable will be the proportion of call backs from each group. Whether or not an applicant has received a callback is found in the variable callback that assigns 0 to applicants who did not receive a callback and a 1 to those who did. Below we can see the outcome variable (the proportion of calls back) for all applicants.

If the proportion of parents who received a callback is smaller than the proportion of non-parents that received a callback then we would see data that supported the thesis on parental discrimination. However if there is little to no difference in proportion then we could have evidence to disprove the hypothesis.

```{r}

library(tidyverse)
library(haven)
library(foreign)
library(readxl)
library(ggplot2)

jobs1 <- read_dta("FieldExp_Public.dta")|>
  drop_na()

jobs <- read_dta("FieldExp_Public.dta")|>
  drop_na()|>
  mutate(parent = if_else(parent==1, "Parent", "Not Parent"), callback =                 if_else(callback==1,"Call back", "No call back"))


prop<-jobs|>
  group_by(callback)|>
  summarize(n=n())|>
  mutate(prop=n/sum(n))

prop|>
  ggplot(aes(x = callback, y = prop, fill = callback)) + geom_col(position = "dodge") +   scale_fill_manual(values = c("Call back"= "green", "No call back" = "red")) +        labs(title="Overall Calls Back", x = "Status", y="Proportions")

knitr::kable(prop, caption = "Proportion of Callbacks", col.names = c("Call Back Status", "Total Number", "Proportion"), digits = 2)
```



Results




According to this barplot, mothers are less likely than fathers to receive a call back if their application signaled they were a mother. Within people who got a callback about 48.8% of them were mothers. In comparison, about 51.2% of them were fathers. Although this is not a huge difference, it prompts further study into what information this data has. I also plan to compare the number of callbacks between women and men as well as parents or non-parents to explain the data shown.


```{r}

parent_prop<-jobs|>
  group_by(parent, callback)|>
  summarize(n=n())|>
  mutate(prop=n/sum(n))

parent_prop|>
  ggplot(aes(x = parent, y = prop, fill = parent)) + geom_col(position = "dodge") + scale_fill_manual(values = c("Parent"= "steelblue1", "Not Parent" = "indianred1")) + labs(title="Percentage of Callbacks between Parents and Non-Parents", x = "Parental Status", y="Proportion of Calls Back")

knitr::kable(parent_prop,  caption = "Proportion of Callbacks between Parents & Non-Parents", col.names = c("Parental Status", "Call Back Status", "Total Number", "Proportion"), digits = 2)


fit <- lm(callback ~ parent, data = jobs1)

fit



```


Mother vs Woman

```{r}
mother_prop<-jobs|>
  filter(woman == 1)|>
  group_by(parent, callback)|>
  summarize(n=n())|>
  mutate(prop=n/sum(n))

mother_prop|>
  ggplot(aes(x = parent, y = prop, fill = parent)) + geom_col(position = "dodge") +      scale_fill_manual(values = c("Parent"= "steelblue1", "Not Parent" = "indianred1"))+    labs(title="Proportion of Callbacks for Women", x = "Parent", y="Proportion of Calls Back")


knitr::kable(mother_prop, caption = "Proportion of Callbacks for Women", col.names = c("Parental Status", "Call Back Status", "Total Number", "Proportion"), digits = 2)
```

Father vs Man
```{r}
father_prop<-jobs|>
  filter(woman == 0)|>
  group_by(parent, callback)|>
  summarize(n=n())|>
  mutate(prop=n/sum(n))

father_prop|>
  ggplot(aes(x = parent, y = prop, fill = parent)) + geom_col(position = "dodge") + scale_fill_manual(values = c("Parent"= "steelblue1", "Not Parent" = "indianred1")) + labs(title="Proportion of Callbacks for Men", x = "Parent", y="Proportion of Calls Back")

knitr::kable(father_prop, caption = "Proportion of Callbacks for Men",col.names = c("Parental Status","Call Back Status",  "Total Number", "Proportion"), digits = 2)
```

```{r}
women_prop<-jobs|>
  mutate(Gender = if_else(woman==1, "Woman", "Man"))|>
  group_by(Gender, callback)|>
  summarize(n=n())|>
  mutate(prop=n/sum(n))

women_prop|>
  ggplot(aes(x = Gender, y = prop, fill = Gender)) + geom_col(position = "dodge") + scale_fill_manual(values = c("Woman"= "steelblue1", "Man" = "indianred1")) + labs(title="Proportion of Callbacks between Genders", x = "Gender", y="Proportion of Calls Back")

knitr::kable(women_prop,  caption = "Proportion of Callbacks between Gender", col.names = c("Gender","Call Back Status",  "Total Number", "Proportion"), digits = 2)
```



